A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-11-25 analysis/paper/paper.Rmd
Minor changes

The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger in 1905 (see @maccurdy1906). Shoreline dating has been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently spatial. Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @crema2010; @crema2012], a similar approach is adopted here.

@@

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet, the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift has been more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable sea-level, been subject to past marine transgression phases, or are still inundated today [e.g. @svendsen1987; @romundset2015]. In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle. Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 210 m above present sea level in the inner Oslo fjord , any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgression phases. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.

@@

Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [e.g. @breivik2018; @solheim2020; @åstveit2018\]. This approach has a couple of limitations. First of all, only a couple of studies have taken the variable uplift-rates into account when performing this comparison. Secondly, following this method, the wider the uncertainty range associated with either radiocarbon- or shoreline date, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. While two overlapping confidence intervals means that one cannot conclude that the dates are different, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no signficant difference to one of equivalence, as the question of interest in this case is effectively one of synchroneity \[cf. @parnell2008]. A final limitation of this method is that it only takes into account the vertical distance between site and shoreline. While this is the parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on local topography and bathymetry.

In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island. In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site, which would then have been accessible by means of watercraft. The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shorebound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which can be guided by a mental or digital reconstruction of the landscape. typically do not produce material suitable for radiocarbon dating. the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.

@@

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the meter distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

@@

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site. This year then has a corresponding likely elevation range for the contemporaneous shoreline. An elevation value was then drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package for R. The topographic distance was measured between site polygon and the point on the shoreline found to be horisontally closest. This was done using In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values where therefore retained.

Some sources of likely varation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, have not been taken into account. Secondly, the DTM has a vertical error of. Thirdly, the tidal range has not been considered. This is around 0.5 m within the study area today, and there is no reason to think this was substantially different in the past. Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.

@@

The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

>>>>

The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in english). Shoreline dating has been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline dating for Norwegian Stone Age archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently spatial. Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @crema2010; @crema2012], a similar approach is adopted here.

@@

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet, the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable relative sea-level, been subject to past marine transgression phases, or remain inundated to this day [e.g. @svendsen1987; @romundset2015]. The has direct archaeological impact. In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle. Transgression, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 210 m above present sea level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time.

@@

Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which can be guided by a both a digital and mental reconstruction of past sea level. Following an excavation, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to assume a shore bound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.

Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island, it would make intuitive sense that the site was used when the ocean reached close to its elevation, which would then have made the site accessible by means of watercraft. The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [e.g. @breivik2018; @solheim2020; @åstveit2018\]. This approach has a couple of limitations. First of all, only a couple of studies have taken the variable uplift-rates into account when performing this comparison \[@fossum2020\]. Secondly, following this method, the wider the uncertainty range associated with either radiocarbon- or shoreline date, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. While two overlapping confidence intervals means that one cannot conclude that the dates are different, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no signficant difference to one of equivalence, as the question of interest in this case is effectively one of synchroneity \[cf. @parnell2008]. Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the shoreline. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on local topography and bathymetry.

In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions.

@@

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the meter distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve. The result of this process is shown for an example site in .

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site. This year then has a corresponding likely elevation range for the contemporaneous shoreline. An elevation value was then drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM. This was done using a topoDist package for R. The topographic distance was measured between site polygon and the point on the shoreline found to be horisontally closest. The shortest path was found using the Queen's case neighbourhood of eight cells. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygon completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values where therefore retained.

Some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, have not been taken into account. Secondly, the DTM has a vertical error of. Thirdly, the tidal range has not been considered. This is around 0.5 m within the study area today, and there is no reason to think this was substantially different in the past. Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.

@@

Here the approach was only used to inform the question of the distance between sites and shorelines, but this method can also be extended to a wide range of use-cases where one needs to visualise as well as quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore line, and the uncertainty inherent in this reconstruction.
